\section{Lessons Learned}\label{section5}

During the execution of the activities documented in this paper, the authors identified some situations in which works related to the concepts involved can benefit. As lessons learned from the work conducted, we highlight:

\begin{itemize}
    \item \textbf{Domain Characteristics:} domain analysis can be considered one of the most important activities for the creation of an SPL. Therefore, the evolution of the catalog of requirements proposed by Duarte Filho and Barbosa \cite{filho13} significantly contributed in terms of domain knowledge. Moreover, it supported the adoption of the proactive model to the development of M-SPLearning.
    \item \textbf{Variabilities and M-Learning:} the use of the SMarty notation helps to identify the variation points during the design of M-SPLearning, ensuring greater cohesion for implementing the SPL components. The notation also contributes to the assimilation of the concept of SPL. However, it is necessary to train those involved so that the elements of notation can be used consistently.
    \item \textbf{M-SPLearning Development:} it is important to notice that implementing something generic and customizable is significantly different of to implement something static. Therefore, developing features in an SPL is considerably more labor intensive, which is justified by the subsequent gains of reuse.
    \item \textbf{M-SPLearning Experimental Evaluation:} researches show that test executions in SPLs are scarce and need to be evaluated and validated \cite{engstrom11}. Thus, the authors decided to apply the tests in the products generated by the SPL, enabling the comparison with an alternative methodology of development.
  
    The experimental evaluation provided relevant results for the adoption of M-SPLearning. The choice for active participants in the industry contributed to the reduction of the training session. However, experience and understanding the concepts by the participants is always difficult to measure, even with the evaluation forms applied. 
\end{itemize}